---
title: "Untitled"
author: "SM"
date: "November 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = "C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/RWJF")
```

```{r}
library(tidyverse)
getwd()
```
Uploading data
```{r}
IV_2010_1 <- read.csv(file="IV_2010_1_csv.csv")
IV_2011_1 <- read.csv(file="IV_2011_1_csv.csv")
IV_2012_1 <- read.csv(file="IV_2012_1_csv.csv")
IV_2013_1 <- read.csv(file="IV_2013_1_csv.csv")
IV_2014_1 <- read.csv(file="IV_2014_1_csv.csv")
IV_2015_1 <- read.csv(file="IV_2015_1_csv.csv")
IV_2016_1 <- read.csv(file="IV_2016_1_csv.csv")
```
Delete columns
```{r}
IV_2010_1 <- IV_2010_1[, -c(38:39, 41:71)]
IV_2012_1 <- IV_2012_1[, -c(40:73)]
IV_2013_1 <- IV_2013_1[, -c(31:61)]
IV_2013_1 <- filter(IV_2013_1, County != "") #delete empty county rows for 2013
IV_2014_1 <- IV_2014_1[, -c(39:81)]
IV_2016_1 <- IV_2016_1[, -c(39:83)]
```

Add year
```{r}
IV_2010_1$year <- NA
IV_2010_1 <- IV_2010_1[,c(1:3,41,4:40)]
IV_2010_1$year <- 2010

IV_2011_1$year <- NA
IV_2011_1 <- IV_2011_1[,c(1:3,39,4:38)]
IV_2011_1$year <- 2011

IV_2012_1$year <- NA
IV_2012_1 <- IV_2012_1[,c(1:3,40,4:39)]
IV_2012_1$year <- 2012

IV_2013_1$year <- NA
IV_2013_1 <- IV_2013_1[,c(1:3,31,4:30)]
IV_2013_1$year <- 2013

IV_2014_1$year <- NA
IV_2014_1 <- IV_2014_1[,c(1:3,39,4:38)]
IV_2014_1$year <- 2014

IV_2015_1$year <- NA
IV_2015_1 <- IV_2015_1[,c(1:3,39,4:38)]
IV_2015_1$year <- 2015

IV_2016_1$year <- NA
IV_2016_1 <- IV_2016_1[,c(1:3,39,4:38)]
IV_2016_1$year <- 2016
```

Initial binding rows
```{r}
IV <- bind_rows(IV_2010_1,IV_2011_1,IV_2012_1,IV_2013_1,IV_2014_1,IV_2015_1,IV_2016_1)
```
Check naming inconsistencies
```{r}
str(IV)
```
Rename 2011 "PCP.Rate"" to "PCP_rate"
```{r}
str(IV_2011_1$PCP.Rate)
names(IV_2011_1)[names(IV_2011_1) == 'PCP.Rate'] <- 'PCP_rate'
```
Rename 2014 "Physically.Inactive" to "inactive_percent"
```{r}
str(IV_2014_1$X..Physically.Inactive)
names(IV_2014_1)[names(IV_2014_1) == 'X..Physically.Inactive'] <- 'inactive_percent'
```
Remove IV dataframe
```{r}
rm(IV)
```
Remake IV dataframe
```{r}
IV <- bind_rows(IV_2010_1,IV_2011_1,IV_2012_1,IV_2013_1,IV_2014_1,IV_2015_1,IV_2016_1)
str(IV)
names(IV)[names(IV) == 'ï..FIPS'] <- 'FIPS'
names(IV)[names(IV) == 'MORT_quart.1'] <- 'MORB_quart'
```

Convert rows to numerics
```{r}
IV$HO_rank <- as.numeric(IV$HO_rank)
IV$HO_quart <- as.numeric(as.character(IV$HO_quart))
IV$HF_rank <- as.numeric(IV$HF_rank)
IV$HF_quart <- as.numeric(as.character(IV$HF_quart))
IV$MORT_rank <- as.numeric(IV$MORT_rank)
IV$MORT_quart <- as.numeric(as.character(IV$MORT_quart))
IV$MORB_rank <- as.numeric(IV$MORB_rank)
IV$MORB_quart <- as.numeric(as.character(IV$MORB_quart))
IV$HB_rank <- as.numeric(IV$HB_rank)
IV$HB_quart <- as.numeric(as.character(IV$HB_quart))
IV$CC_rank <- as.numeric(IV$CC_rank)
IV$CC_quart <- as.numeric(as.character(IV$CC_quart))
IV$SE_rank <- as.numeric(IV$SE_rank)
IV$SE_quart <- as.numeric(as.character(IV$SE_quart))
IV$PE_rank <- as.numeric(IV$PE_rank)
IV$PE_quart <- as.numeric(as.character(IV$PE_quart))
```
Examine structure of IV
```{r}
str(IV)
```

```{r}
setwd("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/DS-Proj_Pull")
save(IV,file="Predictors_1.Rda")
```
Predictors_1-Key
•	N_RankedCounties – no of ranked counties
•	HO_rank – health outcome rank
•	HO_quart – health outcome quartile
•	HF_rank – health factor rank
•	HF_quart – health factor quartile
•	MORT_rank – mortality rank
•	MORT_quart – mortality quartle
•	MORB_rank – morbidity rank
•	MORB_quartile – morbidity quartile
•	HB_rank – health behavior rank
•	HB_quartile – health behavior quartile
•	CC_rank – clinical care rank
•	CC_quart – clinical care quartile
•	SE_rank – social economic factor rank
•	SE_quart – social economic factor quartile
•	PE_rank – physical environment rank
•	PE_quart – physical environment rank
•	Deaths
•	Population
•	YPPL_rate – Years of potential life lost rate
•	Fair_percent – Fair/poor percent
•	PPH_days – poor physical health days
•	PMH_days – poor mental health days 
•	LBW_percent – low birth weight
•	smoking_percent – smoking percent
•	Obese_percent – obese percent
•	inactive_percent – physically inactive percent
•	Binge_percent – binge drinking percent
•	MV_rate – motor vehicle mortality rate
•	CHL_100000 – chlamidia per 100000
•	TeenB_rate – Teen birth rate
•	Uninsure_percent – uninsured percent
•	PCP_rate – primary care provider rate
•	diab_rate – diabetes rate
•	hospice_percent – hospice rate
•	college_percent – college graduate percent
•	unemploy_rate – unemployment
•	Childpov_percent 
