---
title: "Master Data Cleaning"
author: "SM"
date: "December 10, 2017"
output: html_document
---

###Overview and Motivation
We were interested in exploring domestic health policy through the lens of county-level health rankings, mortality data, and other relevant determinants of health. Across the United States, there is broad diversity and variation in health status, access to care, and health outcomes.
We are curious about the decline of life expectancy in the US (for the first time since the 1990s), and which groups of people (by location, race, income level, or other factors) experienced different changes in mortality and by what cause of death. This analysis could provide insights that could guide targeted health interventions for specific communities in the US.
We applied principal component analysis to both Mississippi and Hawaii, comparing to a "healty state" Hawaii. 

###Related Work
As we embarked on this project, we found that we weren’t the only ones interested in exploring disparities in life expectancy/determinants of health at the county level.



###Data

##Independent Variables from the RWJF

Prior to uploading to R, Excel files were downloaded from the RWJF Country Health ranking website for the years 2010-2016. Files were examined in Excel and formatted correctly to make compatible with R.This included deleting rows above the header and identifying andrenaming variables relevant to the analysis.


```{r}
IV_2010_1 <- read.csv(file="IV_2010_1_csv.csv") #Uploading data
IV_2011_1 <- read.csv(file="IV_2011_1_csv.csv")
IV_2012_1 <- read.csv(file="IV_2012_1_csv.csv")
IV_2013_1 <- read.csv(file="IV_2013_1_csv.csv")
IV_2014_1 <- read.csv(file="IV_2014_1_csv.csv")
IV_2015_1 <- read.csv(file="IV_2015_1_csv.csv")
IV_2016_1 <- read.csv(file="IV_2016_1_csv.csv")
IV_2010_1 <- IV_2010_1[, -c(38:39, 41:71)] #Delete columns
IV_2012_1 <- IV_2012_1[, -c(40:73)]
IV_2013_1 <- IV_2013_1[, -c(31:61)]
IV_2013_1 <- filter(IV_2013_1, County != "") #delete empty county rows for 2013
IV_2014_1 <- IV_2014_1[, -c(39:81)]
IV_2016_1 <- IV_2016_1[, -c(39:83)]

IV_2010_1$year <- NA #Add year
IV_2010_1 <- IV_2010_1[,c(1:3,41,4:40)]
IV_2010_1$year <- 2010

IV_2011_1$year <- NA
IV_2011_1 <- IV_2011_1[,c(1:3,39,4:38)]
IV_2011_1$year <- 2011

IV_2012_1$year <- NA
IV_2012_1 <- IV_2012_1[,c(1:3,40,4:39)]
IV_2012_1$year <- 2012

IV_2013_1$year <- NA
IV_2013_1 <- IV_2013_1[,c(1:3,31,4:30)]
IV_2013_1$year <- 2013

IV_2014_1$year <- NA
IV_2014_1 <- IV_2014_1[,c(1:3,39,4:38)]
IV_2014_1$year <- 2014

IV_2015_1$year <- NA
IV_2015_1 <- IV_2015_1[,c(1:3,39,4:38)]
IV_2015_1$year <- 2015

IV_2016_1$year <- NA
IV_2016_1 <- IV_2016_1[,c(1:3,39,4:38)]
IV_2016_1$year <- 2016

IV <- bind_rows(IV_2010_1,IV_2011_1,IV_2012_1,IV_2013_1,IV_2014_1,IV_2015_1,IV_2016_1) #Initial binding rows

str(IV_2011_1$PCP.Rate) #Rename 2011 "PCP.Rate"" to "PCP_rate"

names(IV_2011_1)[names(IV_2011_1) == 'PCP.Rate'] <- 'PCP_rate'

str(IV_2014_1$X..Physically.Inactive)
names(IV_2014_1)[names(IV_2014_1) == 'X..Physically.Inactive'] <- 'inactive_percent' #Rename 2014 "Physically.Inactive" to "inactive_percent"
IV <- bind_rows(IV_2010_1,IV_2011_1,IV_2012_1,IV_2013_1,IV_2014_1,IV_2015_1,IV_2016_1)
str(IV)
names(IV)[names(IV) == 'ï..FIPS'] <- 'FIPS'
names(IV)[names(IV) == 'MORT_quart.1'] <- 'MORB_quart'

#Convert rows to numerics
IV$HO_rank <- as.numeric(IV$HO_rank)
IV$HO_quart <- as.numeric(as.character(IV$HO_quart))
IV$HF_rank <- as.numeric(IV$HF_rank)
IV$HF_quart <- as.numeric(as.character(IV$HF_quart))
IV$MORT_rank <- as.numeric(IV$MORT_rank)
IV$MORT_quart <- as.numeric(as.character(IV$MORT_quart))
IV$MORB_rank <- as.numeric(IV$MORB_rank)
IV$MORB_quart <- as.numeric(as.character(IV$MORB_quart))
IV$HB_rank <- as.numeric(IV$HB_rank)
IV$HB_quart <- as.numeric(as.character(IV$HB_quart))
IV$CC_rank <- as.numeric(IV$CC_rank)
IV$CC_quart <- as.numeric(as.character(IV$CC_quart))
IV$SE_rank <- as.numeric(IV$SE_rank)
IV$SE_quart <- as.numeric(as.character(IV$SE_quart))
IV$PE_rank <- as.numeric(IV$PE_rank)
IV$PE_quart <- as.numeric(as.character(IV$PE_quart))

#Examine structure of IV
str(IV)
#Save
setwd("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/DS-Proj_Pull")
save(IV,file="Predictors_1.Rda")
```

##Z-Scores
The predictors were theh converted to z-scores to standardize the ranks. This was initially done for 2015, but we then changed our analysis to 2014. Therefore the code reflected below specifies 2015 but is working with 2014 data and saved in the 2014 dataframe. 


```{r}
RWJFMeasures_2015 <- read.csv("C:IV_2014_healthrank.csv") #loading 2014 data
str(RWJFMeasures_2015)
names(RWJFMeasures_2015)[names(RWJFMeasures_2015) == 'ï..FIPS'] <- 'FIPS'
RWJFMeasures_2015$HO_rank <- as.numeric(as.character(RWJFMeasures_2015$HO_rank))
RWJFMeasures_2015$HF_rank <- as.numeric(as.character(RWJFMeasures_2015$HF_rank))
RWJFMeasures_2015$MORT_rank <- as.numeric(as.character(RWJFMeasures_2015$MORT_rank))
RWJFMeasures_2015$MORB_rank <- as.numeric(as.character(RWJFMeasures_2015$MORB_rank))
RWJFMeasures_2015$HB_rank <- as.numeric(as.character(RWJFMeasures_2015$HB_rank))
RWJFMeasures_2015$CC_rank <- as.numeric(as.character(RWJFMeasures_2015$CC_rank))
RWJFMeasures_2015$SE_rank <- as.numeric(as.character(RWJFMeasures_2015$SE_rank))
RWJFMeasures_2015$PE_rank <- as.numeric(as.character(RWJFMeasures_2015$PE_rank))
str(RWJFMeasures_2015)
RWJFMeasures_2015$HO_rank <- ave(RWJFMeasures_2015$HO_rank, RWJFMeasures_2015$State, FUN=scale)

#Get zscores
RWJFMeasures_2015$HF_rank <- ave(RWJFMeasures_2015$HF_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$MORT_rank <- ave(RWJFMeasures_2015$MORT_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$MORB_rank <- ave(RWJFMeasures_2015$MORB_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$HB_rank <- ave(RWJFMeasures_2015$HB_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$CC_rank <- ave(RWJFMeasures_2015$CC_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$SE_rank <- ave(RWJFMeasures_2015$SE_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$PE_rank <- ave(RWJFMeasures_2015$PE_rank, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015 <- add_column(RWJFMeasures_2015, deaths_zscore="Deaths", .after = "Deaths")
RWJFMeasures_2015 <- add_column(RWJFMeasures_2015, pop_zscore="Population", .after = "Population")
RWJFMeasures_2015$deaths_zscore <- ave(RWJFMeasures_2015$Deaths, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$pop_zscore <- ave(RWJFMeasures_2015$Population, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$YPPL_rate <- ave(RWJFMeasures_2015$YPPL_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Fair_percent <- ave(RWJFMeasures_2015$Fair_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$PPH_days <- ave(RWJFMeasures_2015$PPH_days, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$PMH_days <- ave(RWJFMeasures_2015$PMH_days, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$LBW_percent <- ave(RWJFMeasures_2015$LBW_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$smoking_percent <- ave(RWJFMeasures_2015$smoking_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Obese_percent <- ave(RWJFMeasures_2015$Obese_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$FoodEnv_Index <- ave(RWJFMeasures_2015$FoodEnv_Index, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$inactive_percent <- ave(RWJFMeasures_2015$inactive_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Exercise_access_percent <- ave(RWJFMeasures_2015$Exercise_access_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Binge_percent <- ave(RWJFMeasures_2015$Binge_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$CHL_100000 <- ave(RWJFMeasures_2015$CHL_100000, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$TeenB_rate <- ave(RWJFMeasures_2015$TeenB_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Uninsure_percent <- ave(RWJFMeasures_2015$Uninsure_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$PCP_rate <- ave(RWJFMeasures_2015$PCP_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$dentist_rate <- ave(RWJFMeasures_2015$dentist_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$MHP_rate <- ave(RWJFMeasures_2015$MHP_rate, RWJFMeasures_2015$State, FUN=scale)

RWJFMeasures_2015$hbac_rate <- ave(RWJFMeasures_2015$hbac_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$mammo_percent <- ave(RWJFMeasures_2015$mammo_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Hsgrad_rate <- ave(RWJFMeasures_2015$Hsgrad_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$college_percent <- ave(RWJFMeasures_2015$college_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$unemploy_rate <- ave(RWJFMeasures_2015$unemploy_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$Childpov_percent <- ave(RWJFMeasures_2015$Childpov_percent, RWJFMeasures_2015$State, FUN=scale)

RWJFMeasures_2015$singleparent_percent <- ave(RWJFMeasures_2015$singleparent_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$social_assoc_rate <- ave(RWJFMeasures_2015$social_assoc_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$violentcrime_rate <- ave(RWJFMeasures_2015$violentcrime_rate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$airpollution <- ave(RWJFMeasures_2015$airpollution, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$drinkingwater_viol <- ave(RWJFMeasures_2015$drinkingwater_viol, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$severehousing_percent <- ave(RWJFMeasures_2015$severehousing_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$driving_alone_percent <- ave(RWJFMeasures_2015$driving_alone_percent, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$longcommute_percent <- ave(RWJFMeasures_2015$longcommute_percent, RWJFMeasures_2015$State, FUN=scale)

RWJFMeasures_2015$acscrate <- ave(RWJFMeasures_2015$acscrate, RWJFMeasures_2015$State, FUN=scale)
RWJFMeasures_2015$injurydeathrate <- ave(RWJFMeasures_2015$injurydeathrate, RWJFMeasures_2015$State, FUN=scale)

#Mistake made - use for 2014 data
RWJFMeasures_2014 <- RWJFMeasures_2015
```


#Finalize z-scores for performing PCA for Mississippi and West Virginia

```{r}

zscores <- RWJFMeasures_2014[c(1:4,22:49)] #subset health factor variables only
names(zscores) # we are missing alcohol driving deaths!

library(readxl) #download from CHR
alcoholdeaths <- read_excel("~/Documents/Fall2017/BST260/FinalProject/ds-proj/DATA CLEANING/RAW DATA/alcoholdeaths.xlsx",
    col_types = c("numeric", "text", "text",
        "numeric"))
View(alcoholdeaths)

alcoholdeaths$alc_drivingdeath <- ave(alcoholdeaths$alc_drivingdeath, alcoholdeaths$State, FUN=scale) #zscore
zscores <- left_join(zscores,alcoholdeaths, by = c("FIPS"="FIPS","State"="State","County"="County")) #join to main data set

```

Filter to needed states
```{r}

zMS_WV <- zscores %>% filter(State== "Mississippi" | State == "West Virginia") # two states with declinine LE

save(zMS_WV, file="Z-scores_MS_WV")

```

Load data frame of factor names to merge during PCA

```{r}
factornames<- read_excel("~/Documents/Fall2017/BST260/FinalProject/ds-proj/DATA CLEANING/RAW DATA/factornames.xlsx", 
    col_types = c("text","text","text"))

save(factornames, file="factornames")

```


##Demographic data
Demographic data was dowloaded from the American Community Survey. In Excel, rows above the header were deleted before importing into R and relevant variables were renamed.

```{r}
#Uploading Race data
Race_2010 <- read.csv("ACS_2010_Race.csv")
Race_2011 <- read.csv("ACS_2011_Race.csv")
Race_2012 <- read.csv("ACS_2012_Race.csv")
Race_2013 <- read.csv("ACS_2013_Race.csv")
Race_2014 <- read.csv("ACS_2014_Race.csv")
Race_2015 <- read.csv("ACS_2015_Race.csv")
Race_2016 <- read.csv("ACS_2016_Race.csv")
#Delete columns
Race_2010 <- Race_2010[, -c(1)]

#Add year
Race_2010$year <- NA
Race_2010$year <- 2010
Race_2010 <- Race_2010[,c(1,2,9,3:8)]

Race_2011$year <- NA
Race_2011$year <- 2011
Race_2011 <- Race_2011[,c(1,2,9,3:8)]

Race_2012$year <- NA
Race_2012$year <- 2012
Race_2012 <- Race_2012[,c(1,2,9,3:8)]

Race_2013$year <- NA
Race_2013$year <- 2013
Race_2013 <- Race_2013[,c(1,2,9,3:8)]

Race_2014$year <- NA
Race_2014$year <- 2014
Race_2014 <- Race_2014[,c(1,2,9,3:8)]

Race_2015$year <- NA
Race_2015$year <- 2015
Race_2015 <- Race_2015[,c(1,2,9,3:8)]

Race_2016$year <- NA
Race_2016$year <- 2016
Race_2016 <- Race_2016[,c(1,2,9,3:8)]

#Initial binding rows
Race <- bind_rows(Race_2010,Race_2011,Race_2012,Race_2013,Race_2014,Race_2015,Race_2016)

#Check naming inconsistencies
str(Race)
#Delete Race

rm(Race)
#Rename Geography in Race_2015
names(Race_2015)[names(Race_2015) == 'Geography'] <- 'County_State'

#Binding rows

Race <- bind_rows(Race_2010,Race_2011,Race_2012,Race_2013,Race_2014,Race_2015,Race_2016)
str(Race)

#Save data
setwd("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/DS-Proj_Save_SM")
save(Race, file="Race.Rda")
```

Uploading Age data
```{r}
Age_2010 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2010_Age.csv")
Age_2011 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2011_Age.csv")
Age_2012 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2012_Age.csv")
Age_2013 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2013_Age.csv")
Age_2014 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2014_Age.csv")
Age_2015 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2015_Age.csv")
Age_2016 <- read.csv("C:/Users/Sohini Mukherjee/OneDrive/Documents/Harvard University/Academic/Fall 2017/BST 260 - Data Science I/Final Project/ACS/Age/ACS_2016_Age.csv")
```

Add year
```{r}
Age_2010$year <- NA
Age_2010$year <- 2010
Age_2010 <- Age_2010[,c(1,2,89,3:7)]

Age_2011$year <- NA
Age_2011$year <- 2011
Age_2011 <- Age_2011[,c(1,2,91, 3:7)]

Age_2012$year <- NA
Age_2012$year <- 2012
Age_2012 <- Age_2012[,c(1,2,7, 3:6)]

Age_2013$year <- NA
Age_2013$year <- 2013
Age_2013 <- Age_2013[,c(1,2,13, 3:12)]

Age_2014$year <- NA
Age_2014$year <- 2014
Age_2014 <- Age_2014[,c(1,2,12, 3:11)]

Age_2015$year <- NA
Age_2015$year <- 2015
Age_2015 <- Age_2015[,c(1,2,11,3:10)]

Age_2016$year <- NA
Age_2016$year <- 2016
Age_2016 <- Age_2016[,c(1,2,168,3:167)]
```


Choose columns:
```{r}
Age_2013 <- select(Age_2013, FIPS, County_State, year, Total, Male, Female, Under_5, Above_65)
Age_2014<- select(Age_2014, FIPS, County_State, year, Total, Male, Female, Under_5, Above_65)
Age_2015 <- select(Age_2015, FIPS, County_State, year, Male, Female, Under_5, Above_65)
Age_2016 <- select(Age_2016, FIPS, Geography,year, Total, Male, Female, Under_5, Above_65)

```

Fix columns:
```{r}
names(Age_2016)[names(Age_2016) == "Geography"] <- "County_State"
Age_2012$Above_65 <- NA
Age_2015$Total <- Age_2015$Male + Age_2015$Female
```

```{r}
Age <- bind_rows(Age_2010, Age_2011, Age_2012, Age_2013, Age_2014, Age_2015, Age_2016)
str(Age)
```
Save data
```{r}
save(Age, file="Pop_Sex_Age.Rda")
```




